---
title: "Reading & Writing Data in R"
author: "YOUR NAME HERE"
date: 'Last rendered on `r format(Sys.time(), "%A, %B %d, %Y @ %I:%M %p")`'
output: 
  html_document: 
    theme: yeti
    highlight: textmate
    toc: true
    toc_depth: 4
---

```{r globaloptions, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = NA
)
```

```{r}
library(readxl)
library(jsonlite)
```

## RDS and RDATA Files

### Reading and Writing `.rds`

Read an object from an `.rds` file using the following general code. You may need to specify a path name. You can also read data directly from a URL.

```{r, readrds, eval = FALSE}
readRDS(file = "my_data.rds")
```

Save a single object to an `.rds` file using the following general code. If the data are being saved to a folder you need to specify the path to the folder.

```{r, saverds, eval = FALSE}
saveRDS(objectname, file = "my_data.rds")
```

### Reading and Writing `.Rdata`

Load an `.Rdata` file using the following general code. You may need to specify a path name. You can also read data directly from a URL.

```{r, loadrata, eval = FALSE}
load("data.RData")
```

    # syntax
    load(file, 
    envir = parent.frame(), 
    verbose = FALSE)

Save a one or more objects to an `.Rdata` file using the following general code. If the data are being saved to a folder you need to specify the path to the folder.

```{r, saverdata1, eval = FALSE}
save(objectname, file = "data.RData")
```

```{r, saverdata2, eval = FALSE}
save(objectname1, objectname2, file = "data.RData")
```

    # syntax
    save(..., 
      list = character(),
      file = stop("'file' must be specified"),
      ascii = FALSE, 
      version = NULL, 
      envir = parent.frame(),
      compress = isTRUE(!ascii), 
      compression_level,
      eval.promises = TRUE, 
      precheck = TRUE)

Save your entire workspace to an `.Rdata` file using the following general code. If the data are being saved to a folder you need to specify the path to the folder.

```{r, savewkspc, eval = FALSE}
save.image()
```

    # syntax
    save.image(file = ".RData", 
      version = NULL, 
      ascii = FALSE,
      compress = !ascii, 
      safe = TRUE)

## Excel Files

TBA

## JSON Files

TBA